\vspace{-2em} Hamiltonian Privilege

### 2021-07-26 Articles/2020-ClassicalMechanicsPhil/Submissions/BJPS/classical-phil-article-BJPS.tex
Update classical-phil-article-BJPS.tex
Addressed most referee comments; some outstanding `to do's', especially in section 4.2.

Subsequent sections describe additional advantages of Hamiltonian mechanics, along with further interpretive benefits of our approach. Section~\ref{density} explains how the invariant phase space density plays a special role in the Hamiltonian framework. The density allows us to compare and track the amount of matter in different regions of phase space. In contrast, Lagrangian mechanics does not support these important properties of the density. Finally, Section~\ref{Lagrangian} considers the status of Lagrangian mechanics within our framework. We provide a geometric interpretation of the action principle, which applies even outside the hyper-regular domain. We end by rebutting objections that arise from Curiel's \parencites*[]{Curiel} privileging of Lagrangian mechanics. 
%Along the way, we examine the gauge redundancy of both the conjugate momentum (Section~\ref{gauge_momentum}) and the Lagrangian (Section~\ref{gauge}).

@@

In claiming that Hamiltonian mechanics is more fundamental than Lagrangian mechanics, we must clarify what we mean by `fundamental.' Of the various notions of fundamentality, we will mean something entirely non-mysterious: a formalism is more fundamental provided that it is logically more general. In this case, the Lagrangian framework is less general because it requires an additional condition to hold, compared to the Hamiltonian framework. Hence, the Lagrangian framework is less fundamental because it is a proper subset of the Hamiltonian framework.\footnote{One might worry that our criterion for fundamentality is subject to elementary counterexamples. For instance, can't we make any theory more general by adjoining arbitrary disjunctions? Importantly, we require that the theory remains wholly \textit{about} classical mechanics. Hamiltonian mechanics is more general than Lagrangian mechanics while still being about classical systems, where `aboutness' can be understood using Yablo's framework \parencites*[]{Yablo}. Having an account of \textit{aboutness} or \textit{subject matter} is a general philosophical problem that any account of fundamentality or essence must address. Hence, it is not a special problem facing our criterion.}

Our argument has three main steps, summarized in Figure~\ref{diagram}. The left column lists three basic physical conditions: infinitesimal reducibility, determinism/reversibility, and motion equivalence. Each condition entails physically significant properties, described in the middle column. Each physical property corresponds to a formal/mathematical structure depicted in the right column, including phase space, Hamiltonian mechanics, and Lagrangian mechanics. In this way, our framework develops a correspondence between physical conditions and their associated mathematical formalisms. We thereby gain a straightforward interpretation of the physical significance of these formalisms.

First, a system is \textit{infinitesimally reducible} provided that the state of the whole system is equivalent to specifying the states of its parts (and vice versa). Infinitesimal reducibility distinguishes classical from quantum systems, where the latter are irreducible (i.e. we cannot specify the whole state by specifying the states of its subsystems). In the widest sense, a classical system is any infinitesimally reducible system. Colloquially, we can say that the whole state equals the sum of its parts, where we take these infinitesimal parts to be classical particles.

If a system is infinitesimally reducible, then we can define a distribution $\rho$ over particle states and a density $\rho_{\xi}$ over the corresponding state variables $\xi$. Since particle states are independent of a choice of units, this density must be invariant. We show in Section~\ref{infinitesimal} that defining this invariant density requires a symplectic manifold, which serves as the phase space of classical mechanics. The physical significance of this symplectic manifold is thereby its connection with the physical condition of infinitesimal reducibility.

Secondly, a system has \textit{deterministic and reversible evolution} provided that the present state of the system uniquely determines both its future and past states. This condition entails that the invariant density $\rho_{\xi}$ is a conserved phase space density. We will show that a classical system which satisfies this condition is Hamiltonian: its phase space is a symplectic manifold that evolves according to Hamiltonian evolution  (i.e. according to a symplectomorphism). We thereby characterize Hamiltonian mechanics as the mathematical framework for infinitesimally reducible systems with deterministic and reversible evolution. Section~\ref{deterministic} justifies this correspondence.

Finally, some classical systems satisfy a further condition, which we call \textit{motion equivalence}. Motion equivalence holds when the system's trajectory in physical space determines---and is determined by---its evolution in phase space. In other words, knowledge of the trajectory in physical space (e.g. Euclidean three-space $\mathbb{R}^3$) suffices for knowledge of the trajectory in phase space, and vice versa. Physically, knowledge of the trajectory in physical space corresponds to knowledge of position and velocity, while knowledge of the trajectory in phase space requires knowing the system's momentum. Hence, motion equivalence entails a correspondence between velocity and momentum, which formally leads to the framework of Lagrangian mechanics in the hyper-regular domain (Section~\ref{motion}). As a result, Lagrangian mechanics requires that a further physical condition is satisfied, and hence it is less general than Hamiltonian mechanics.

We cannot always determine the evolution of a system in phase space from its trajectory in physical space, since some systems fail to satisfy motion equivalence. In general, knowing a system's position and velocity does not suffice for knowing its momentum. Considering a photon as a classical particle, we cannot determine its momentum from its trajectory through space.\footnote{Since photons travel at constant speed, we can recover the direction of momentum---but not the magnitude---from the trajectory. Two photons with different momenta can follow the same trajectory in physical space.} Hence, knowledge of the physical trajectory of a photon is insufficient for knowledge of its phase space evolution. Moreover, different Lagrangians---differing even by more than a constant---can correspond to the same particle trajectories while not agreeing on the total energy, even up to a constant.\footnote{\textcites[1185--1186]{Barrett2} discusses this in the context of the failure of equivalence between the vector field formulations of Lagrangian and Hamiltonian mechanics. In contrast, the trajectories in phase space (encoded by the Hamiltonian vector field) uniquely define the Hamiltonian up to a constant.} Hence, the set of classical systems that satisfy motion equivalence is a proper subset of the classical systems that satisfy the conditions for Hamiltonian mechanics. For at least this reason, Hamiltonian mechanics is more fundamental than Lagrangian mechanics.

Of course, different definitions of `Hamiltonian' and `Lagrangian' systems can result in different interpretive relationships. On our account, Hamiltonian systems are infinitesimally reducible and follow deterministic and reversible evolution.\footnote{Although not sufficient for a definition, Newtonian systems satisfy the motion equivalence assumption. Since Lagrangian systems satisfy all three conditions (infinitesimal reducibility, deterministic/reversible evolution, and motion equivalence), they constitute the intersection of Newtonian and Hamiltonian systems.} We thereby set aside well-known cases of non-deterministic classical systems.\footnote{For discussion of non-deterministic classical systems, see \textcites*[3-4]{Baez}{Earman}{Norton}.} Additionally, we set aside dissipative systems, i.e. those that fail to conserve energy. Possessing a deterministic and reversible evolution is equivalent to being non-dissipative. Our derivation of Lagrangian mechanics in Section~\ref{motion} takes place within the context of conservative systems, so we do not consider systems that satisfy inhomogenous Euler--Lagrange equations. Although both the Lagrangian and Hamiltonian formulations apply to dissipative systems, we exclude these.\footnote{For details, see \textcites[\S 10.4]{Cline}. Curiel \parencites*[311]{Curiel} acknowledges that the Hamiltonian framework can be extended to treat dissipative systems, although he seems to view it as a virtue of the Lagrangian formulation that it handles dissipative systems with fewer modifications. For details of the Lagrangian approach to dissipative systems, see \textcites[]{Smith}.} This restricts our claim about the fundamentality of Hamiltonian mechanics to the non-dissipative regime. Finally, our treatment is not restricted to $n$-particle mechanics but applies as well to rigid-body and continuum mechanics. This supports our claim that the property of infinitesimal reducibility fruitfully characterizes classical systems. Before presenting the derivational details of our argument in Section~\ref{derivation}, we describe how we avoid central problems facing North's privileging of Hamiltonian mechanics. 
%We view non-dissipative systems as neither Lagrangian nor Hamiltonian. 
%\footnote{For discussion of the numerous conceptual problems that arise within different approaches to classical mechanics (including point-masses, rigid-bodies, and continua) see \textcites{Wilson}. In Wilson's terminology, the framework we apply here may be yet another `theory facade' for classical physics, but a fruitful one nonetheless.}

@@

As indicated at the start of Section~\ref{privileging}, we use `fundamentality' in a non-metaphysical fashion. Some differences in fundamentality are differences in generality: more fundamental frameworks require fewer conditions to hold.\footnote{We also embrace related non-metaphysical uses of `fundamental,' discussed in Section~\ref{density}.} Our notion of fundamentality leads to a correspondingly non-metaphysical notion of `structure.' Hamiltonian mechanics employs less structure than Lagrangian mechanics because the former requires fewer physical conditions. Compared with North's \parencites*[]{North} account of structure, our definition has two advantages. First, it avoids metaphysical disputes surrounding perfectly natural or joint-carving properties, which scientific anti-realists and pragmatists decry. Secondly, it avoids a technical objection that Swanson and Halvorson \parencites*[]{Swanson} levy against North's more precise mathematical criterion for comparing structure.

@@

Additionally, whereas North \parencites*[76]{North} seeks to determine the `real' fundamental state space structure of the theory---and correspondingly, of a classical world---we remain neutral on this underlying metaphysical question.\footnote{Indeed, Swanson and Halvorson \parencites*[]{Swanson} note that many physicists and philosophers would not interpret statespace structure ontologically, but as a formal device for encoding physical information or a calculational tool. As Wilson illustrates, we can interpret classical mechanics while avoiding---and even rejecting---the search for what is `fundamental in the bottom-layer sense' \parencites*[53]{Wilson}.} Whether Hamiltonian phase space structure is metaphysically fundamental in a classical world does not matter for the questions we seek to answer. Regardless of its metaphysical status, Hamiltonian mechanics is physically and conceptually more fundamental. Indeed, our argument does not privilege any one of the various mathematical formalizations of Hamiltonian mechanics. These include formulating Hamiltonian mechanics on a $2n+1$-dimensional contact manifold, on an extended phase space, on a Poisson manifold, and the standard formulation on a symplectic manifold that we recover in Section~\ref{deterministic}.\footnote{Tulczyjew's reformulation of Hamiltonian mechanics further supports viewing symplectic geometry---and hence Hamiltonian mechanics---as conceptually fundamental. According to Teh and Tsementzis, this reformulation shows that  ``the very concept of a classical mechanical system (and not merely the phase space of that system) should be described in terms of symplectic geometry'' \parencites*[46]{Teh}.}

@@

Advantageously, our criterion for comparing Hamiltonian and Lagrangian mechanics avoids these problems. Hamiltonian mechanics has less structure than Lagrangian mechanics simply in the sense of logical strength: a system must satisfy strictly fewer physical conditions in order to be Hamiltonian. Whereas to be Lagrangian, the system must also satisfy the motion equivalence condition. Motion equivalence places an additional constraint on the Hamiltonian, a constraint that need not be satisfied in general. Section~\ref{density} provides a further argument for viewing Hamiltonian mechanics as more fundamental. We argue that (i) invariant phase space densities are physically fundamental and (ii) Hamiltonian mechanics provides the proper setting for representing these densities.

@@

This section justifies the physical and mathematical relationships in Figure~\ref{diagram}.
Sections~\ref{infinitesimal}-\ref{deterministic} derive the formalism of Hamiltonian mechanics from the assumptions of infinitesimal reducibility and deterministic/reversible evolution. Section~\ref{motion} shows how systems that satisfy the further condition of motion equivalence can be treated within the formalism of Lagrangian mechanics. We thus demonstrate in detail our methodology of deriving  relevant mathematical structures from underlying physical conditions.

Before starting, we want to make it clear that the following is not and cannot be a formal proof. The idea is that the assumptions put constraints on the physical entities and, in turn, on the mathematical objects that are used to represent them. These constraints are constitutive in nature, they provide requirements for the math, and therefore they cannot be derived from the math. Each assumption will introduce multiple constraints, which will be introduced gradually within each section. The goal is to show how such constraints follow logically from the assumptions, how they are encoded in the math and how they recover the well known structures.

@@

First, we will show that a classical phase space arises if a system satisfies the following physical condition:

@@

Let $\mathcal{C}$ be the state space for the whole system, so that any point $c \in \mathcal{C}$ represents a particular state of the whole system. Given infinitesimal reducibility, a classical system consists of a collection of classical particles. Let $\mathcal{S}$ be the state space for the particles, so that each point $s \in \mathcal{S}$ describes a possible state of a classical particle. Then for each state $c \in \mathcal{C}$ there exists a unique distribution $\rho_c : \mathcal{S} \to \mathbb{R} $ that describes the state of $c$'s parts. That is, the state of the whole system is a distribution over its infinitesimal parts. This distribution describes the proportion of classical particles in any particular state from $\mathcal{S}$.

@@

When the system is characterized by continuous quantities, $\mathcal{S}$ is a manifold, and we define $\mu(U) = \int_U \rho_c(s) d\mathcal{S} $. State variables $\xi^a$ are any set of quantities that fully identify states, so that $\xi$: $\mathcal{S} \to \mathbb{R}^n$ is an invertible function between each state and the numeric values of the variables.\footnote{In the language of manifolds, these $\xi^a$ are called `coordinates' and $\xi$ is a coordinate map. We will have multiple manifolds, so to avoid confusion, we will call the coordinates of the state space `state variables'. As usual with coordinate charts, $\xi$ is properly defined on some $U \subseteq \mathcal{S}$, and $\xi^{-1}$ is defined on $\xi(U) \subseteq \mathbb{R}^n$. We will neglect these subtleties in what follows.} The map $\xi$ represents a state $s \in \mathcal{S}$ by a set of state variables $\xi^a \in \mathbb{R}^n$, such that $\xi (s) = \xi^a$. To represent the distribution $\rho_c$ in terms of state variables, we define a density $\rho_{c, \xi}$: $\mathbb{R}^n \to \mathbb{R}$ by $\rho_{c, \xi} \equiv (\rho_c \circ \xi^{-1})$. This density plays a crucial role throughout our argument.

Since $\rho_c$: $\mathcal{S} \to \mathbb{R}$ is a function of the particle state $s$, its value is independent of the choice of state variables $\xi^a$ (i.e. the choice of state variable map $\xi$). Hence, the density $\rho_{c, \xi}$ transforms as a scalar under state variable changes.\footnote{For two state variable maps $\xi$ and $\hat{\xi}$, we have $\rho_c(s) = (\rho_c \circ \xi^{-1} \circ \xi)(s) = (\rho_c \circ \hat{\xi}^{-1} \circ \hat{\xi})(s)$, which entails that $\rho_c(s) = \rho_{c, \xi} (\xi^a) = \rho_{c, \hat{\xi}} (\hat{\xi}^b)$. Here, $\xi(s) \equiv \xi^a$ and $\hat{\xi}(s) \equiv \hat{\xi}^b$.} Likewise, the integral $\int_U \rho_c(s) d\mathcal{S} $ is independent of the choice of state variables, so $\int_{\xi(U)} \rho_{c, \xi} (\xi^a) d \xi^a = \int_{\hat{\xi}(U)} \rho_{c, \hat{\xi}} (\hat{\xi}^b) d \hat{\xi}^b$. By the change of variables formula, we also know that $\int_{\hat{\xi}(U)} \rho_{c, \hat{\xi}} (\hat{\xi}^b) d \hat{\xi}^b= \int_{\xi(U)} \rho_{c, \hat{\xi}} (\hat{\xi}^b) \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right|  d \xi^a$.\footnote{The change of variables transformation from $\xi^a$ to $\hat{\xi}^b$ is given by $\phi \equiv \hat{\xi} \circ \xi^{-1}$: $\mathbb{R}^n \to \mathbb{R}^n $. For convenience, we denote $\rho_{c, \hat{\xi}} (\phi(\xi^a))$ by $\rho_{c, \hat{\xi}} (\hat{\xi}^b)$. Likewise, in the numerator of the Jacobian, we denote $\partial \phi$ by $\partial \hat{\xi}^b$, where $\hat{\xi}^b$ is here understood as a function of $\xi^a$.} Here, $ \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right|$ is the determinant of the Jacobian for this change of state variables. Matching integrands from the preceding two equations, we have $\rho_{c, \xi} (\xi^a) = \rho_{c, \hat{\xi}} (\hat{\xi}^b) \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right|$. From the scalar transformation property, $\rho_{c, \xi} (\xi^a) = \rho_{c, \hat{\xi}} (\hat{\xi}^b)$. Hence, we see that $\rho_{c, \hat{\xi}} (\hat{\xi}^b) =  \rho_{c, \hat{\xi}} (\hat{\xi}^b) \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right|$, entailing that the Jacobian determinant must equal one. By dimensional homogeneity, it must also be dimensionless. We will now show that in order for the density $\rho_{c, \xi}$ to have these transformation properties, the classical phase space must be at least a symplectic manifold. 
%From the preceding two equations, we see that $\int_{\xi(U)} \rho_{c, \xi} (\xi^a) d \xi^a = \int_{\xi(U)} \rho_{c, \hat{\xi}} (\hat{\xi}^b) \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right|  d \xi^a$.

>>>>

Subsequent sections describe additional advantages of Hamiltonian mechanics, along with further interpretive benefits of our approach. Section~\ref{density} explains how the invariant phase space density plays a special role in the Hamiltonian framework. The density allows us to compare and track the amount of matter in different possible states. In contrast, Lagrangian mechanics does not support these important properties of the density. Finally, Section~\ref{Lagrangian} considers the status of Lagrangian mechanics within our framework. We provide a geometric interpretation of the action principle, which applies even outside the hyper-regular domain. We end by rebutting objections that arise from Curiel's \parencites*[]{Curiel} privileging of Lagrangian mechanics. 
%Along the way, we examine the gauge redundancy of both the conjugate momentum (Section~\ref{gauge_momentum}) and the Lagrangian (Section~\ref{gauge}).

@@

In claiming that Hamiltonian mechanics is more fundamental than Lagrangian mechanics, we must clarify what we mean by `fundamental.' Of various notions of fundamentality, we mean something entirely non-mysterious. Given two formalisms that capture a class of systems, if one is logically more general, then it is conceptually more fundamental.\footnote{In Section~\ref{density}, we embrace a related non-metaphysical notion of fundamentality that pertains to physical entities rather than theoretical frameworks.} Conceptual fundamentality is familiar from mathematics. A more fundamental framework encompasses a less fundamental one as a special case. It is in this sense that the generalized Stokes theorem is more fundamental than well-known special cases, such as the fundamental theorem of calculus and Green's theorem. Similarly, an axiomatization which includes additional axioms is less fundamental than one which drops these axioms while maintaining the same expressive power. Conceptual fundamentality is distinct from notions of fundamentality associated with differences in energy scale, such as how the Standard Model of particle physics is more fundamental than classical mechanics. We do not mean that Hamiltonian mechanics is more fundamental than Lagrangian mechanics in this sense.

To apply our criterion for fundamentality, we must show that Lagrangian mechanics is less general than Hamiltonian mechanics. We do this in Section~\ref{derivation} by recovering the mathematical formalisms of both Hamiltonian and Lagrangian mechanics from three physical conditions. Our aim here is to show how this argument works in principle. Section~\ref{fundamentality} demonstrates some philosophical advantages of our preferred notion of fundamentality.\footnote{One might worry that our criterion for fundamentality is subject to elementary counterexamples. For instance, can't we make any theory more general by adjoining arbitrary disjunctions? Importantly, we require that the theory remains wholly \textit{about} classical mechanics. Hamiltonian mechanics is more general than Lagrangian mechanics while still being about classical systems, where `aboutness' can be understood using Yablo's framework \parencites*[]{Yablo}. Having an account of \textit{aboutness} or \textit{subject matter} is a general philosophical problem that any account of fundamentality must address. It is not a special problem facing our criterion.} 
%We will show that the Lagrangian framework is less general because it requires an additional condition to hold, compared to the Hamiltonian framework. Hence, the Lagrangian framework is less fundamental because it is a proper subset of the Hamiltonian framework.

%We will, however in Section 5 give a criterion for an entity to be physically fundamental
%These bolster our claim that this is an appropriate sense of fundamentality for philosophers to focus on.

Shorn of many details, our argument goes as follows. First, we show that the formalism of Hamiltonian mechanics applies to systems that are both i) infinitesimally reducible and ii) follow a deterministic and reversible evolution. A system is \textit{infinitesimally reducible} provided that the state of the whole system is equivalent to specifying the states of its parts (and vice versa). Colloquially, we can say that the whole state equals the sum of its parts, where we take these infinitesimal parts to be classical particles. Infinitesimal reducibility distinguishes classical from quantum systems, where the latter are irreducible (i.e. we cannot specify the whole state by specifying the states of its subsystems). We show in Section~\ref{infinitesimal} that the phase space of infinitesimally reducible systems must have at least the structure of a symplectic manifold. Recovering Hamiltonian mechanics then simply requires identifying when trajectories on phase space obey Hamiltonian evolution  (i.e. evolve according to a symplectomorphism). Section~\ref{deterministic} shows that this requires the system to follow a \textit{deterministic and reversible evolution}, i.e. the present state of the system uniquely determines both its future and past states. We thereby characterize Hamiltonian mechanics as the mathematical framework for infinitesimally reducible systems with deterministic and reversible evolution.

Next, we note that some classical systems satisfy a third, further condition, which we call \textit{motion equivalence}. Motion equivalence holds when the system's trajectory in physical space determines---and is determined by---its evolution in phase space. In other words, knowledge of the trajectory in physical space (e.g. Euclidean three-space $\mathbb{R}^3$) suffices for knowledge of the trajectory in phase space, and vice versa. Physically, knowledge of the trajectory in physical space corresponds to knowledge of position and velocity, while knowledge of the trajectory in phase space requires knowing the system's momentum. Hence, motion equivalence entails a correspondence between velocity and momentum, which formally leads to the framework of Lagrangian mechanics in the hyper-regular domain (Section~\ref{motion}). As a result, Lagrangian mechanics requires that a further physical condition is satisfied, and hence it is less general than Hamiltonian mechanics. 
% Argument here could be logically more clear: note that we next have to recover Lagrangian mechanics and do so through an additional condition.

To show that motion equivalence is a further condition, it suffices to show that some Hamiltonian systems violate it. This follows from the fact that knowing a system's position and velocity does not always suffice for knowing its momentum. Hence, we cannot always determine the evolution of a system in phase space from its trajectory in physical space. For instance, if we consider a photon as a classical particle, then we cannot determine its momentum from its trajectory through space. Since photons travel at constant speed, we can recover the direction of momentum---but not the magnitude---from the trajectory. Two photons with different momenta can follow the same trajectory in physical space. Hence, knowledge of the physical trajectory of a photon is insufficient for knowledge of its phase space evolution. In general, different Lagrangians---differing even by more than a constant---can correspond to the same particle trajectories while not agreeing on the total energy, even up to a constant.\footnote{\textcites[1185--1186]{Barrett2} discusses this in the context of the failure of equivalence between the vector field formulations of Lagrangian and Hamiltonian mechanics. In contrast, the trajectories in phase space (encoded by the Hamiltonian vector field) uniquely define the Hamiltonian up to a constant.} Hence, the set of classical systems that satisfy motion equivalence is a proper subset of the classical systems that satisfy the conditions for Hamiltonian mechanics. For at least this reason, Hamiltonian mechanics is more fundamental than Lagrangian mechanics.

% % To do: address referee 1 concern about the clarity of this argument/example: ``what if momentum is defined in terms of velocity? Or is the idea to know that we also need to know the mass?"

Of course, different definitions of `Hamiltonian' and `Lagrangian' systems can result in different interpretive relationships. On our account, Hamiltonian systems are infinitesimally reducible and follow deterministic and reversible evolution.\footnote{Although not sufficient for a definition, Newtonian systems also satisfy the motion equivalence assumption. Since Lagrangian systems satisfy all three conditions (infinitesimal reducibility, deterministic/reversible evolution, and motion equivalence), they lie in the intersection of Newtonian and Hamiltonian systems.} We thereby set aside well-known cases of non-deterministic classical systems.\footnote{For discussion of non-deterministic classical systems, see \textcites*[3-4]{Baez}{Earman}{Norton}.} Additionally, we set aside dissipative systems, i.e. those that fail to conserve energy. Possessing a deterministic and reversible evolution is equivalent to being non-dissipative. Our derivation of Lagrangian mechanics in Section~\ref{motion} takes place within the context of conservative systems, so we do not consider systems that satisfy inhomogenous Euler--Lagrange equations. Although both the Lagrangian and Hamiltonian formulations apply to dissipative systems, we exclude these.\footnote{For details, see \textcites[\S 10.4]{Cline}. Curiel \parencites*[311]{Curiel} acknowledges that the Hamiltonian framework can be extended to treat dissipative systems, although he seems to view it as a virtue of the Lagrangian formulation that it handles dissipative systems with fewer modifications. For details of the Lagrangian approach to dissipative systems, see \textcites[]{Smith}.} This restricts our claim about the fundamentality of Hamiltonian mechanics to the non-dissipative regime. Finally, our treatment is not restricted to $n$-particle mechanics but applies as well to rigid-body and continuum mechanics. This supports our claim that the property of infinitesimal reducibility fruitfully characterizes classical systems. Before presenting the details of our argument in Section~\ref{derivation}, we describe how we avoid central problems facing North's privileging of Hamiltonian mechanics.

@@

As indicated at the start of Section~\ref{privileging}, we use `fundamentality' in a non-metaphysical fashion. Some differences in fundamentality are differences in generality: provided a theoretical framework still captures the relevant behavior, it is more fundamental if it requires fewer conditions to hold. Our notion of fundamentality leads to a correspondingly non-metaphysical notion of `structure.' Hamiltonian mechanics employs less structure than Lagrangian mechanics because the former requires fewer physical conditions. Compared with North's \parencites*[]{North} account of structure, our definition has two advantages. First, it avoids metaphysical disputes surrounding perfectly natural or joint-carving properties, which scientific anti-realists and pragmatists decry. Secondly, it avoids a technical objection that Swanson and Halvorson \parencites*[]{Swanson} levy against North's more precise mathematical criterion for comparing structure.

@@

Additionally, whereas North \parencites*[76]{North} seeks to determine the `real' fundamental state space structure of the theory---and correspondingly, of a classical world---we remain neutral on this underlying metaphysical question.\footnote{Indeed, Swanson and Halvorson \parencites*[]{Swanson} note that many physicists and philosophers would not interpret statespace structure ontologically, but as a formal device for encoding physical information or a calculational tool. As Wilson illustrates, we can interpret classical mechanics while avoiding---and even rejecting---the search for what is `fundamental in the bottom-layer sense' \parencites*[53]{Wilson}.} Whether Hamiltonian phase space structure is metaphysically fundamental in a classical world does not matter for the questions we seek to answer. Regardless of its metaphysical status, Hamiltonian mechanics is conceptually more fundamental. Indeed, our argument does not privilege any one of the various mathematical formalizations of Hamiltonian mechanics. These include formulating Hamiltonian mechanics on a $2n+1$-dimensional contact manifold, on an extended phase space, on a Poisson manifold, and the standard formulation on a symplectic manifold that we recover in Section~\ref{deterministic}.\footnote{Tulczyjew's reformulation of Hamiltonian mechanics further supports viewing symplectic geometry---and hence Hamiltonian mechanics---as conceptually fundamental. According to Teh and Tsementzis, this reformulation shows that  ``the very concept of a classical mechanical system (and not merely the phase space of that system) should be described in terms of symplectic geometry'' \parencites*[46]{Teh}.}

@@

Advantageously, our criterion for comparing Hamiltonian and Lagrangian mechanics avoids these problems. Hamiltonian mechanics has less structure than Lagrangian mechanics simply in the sense of logical strength: a system must satisfy strictly fewer physical conditions in order to be Hamiltonian. Whereas to be Lagrangian, the system must also satisfy the motion equivalence condition. Motion equivalence places an additional constraint on the Hamiltonian, a constraint that need not be satisfied in general. Section~\ref{density} provides a further argument for viewing Hamiltonian mechanics as more fundamental. We argue that (i) invariant phase space densities are physically fundamental and (ii) Hamiltonian mechanics provides the proper setting for representing these densities.
%is this the actual phase space density vs. density for all possible trajectories?

@@

% % try to shorten intro to this section, since parts are repetitive relative to section 2

Our derivation has three main steps. Section~\ref{infinitesimal} shows that the phase space of infinitesimally reducible systems must have at least the structure of a symplectic manifold. For systems with deterministic and reversible evolution, Section~\ref{deterministic} shows they must evolve according to a symplectomorphism. Hence, we derive the formalism of Hamiltonian mechanics from the conditions of infinitesimal reducibility and deterministic/reversible evolution. Section~\ref{motion} shows how systems that satisfy the further condition of motion equivalence can be treated within the formalism of Lagrangian mechanics.

%-\ref{deterministic} derive the formalism of Hamiltonian mechanics from the assumptions of infinitesimal reducibility and deterministic/reversible evolution. along with a few minor assumptions introduced below

Figure~\ref{diagram} summarizes the results of this derivation. The left column lists the three central physical conditions. Each condition entails physically significant properties, described in the middle column. Each physical property corresponds to a formal/mathematical structure depicted in the right column, including phase space, Hamiltonian mechanics, and Lagrangian mechanics. In this way, our framework develops a correspondence between physical conditions and their associated mathematical formalisms. We thereby elucidate the physical significance of these formalisms.
%details of this figure will become clear as we go along.

Admittedly, our derivation falls short of some norms on mathematical rigor. We use the three physical conditions to place constraints on physical entities and, in turn, on the mathematical objects that can be used to represent them. Each condition leads to additional constraints, introduced gradually within each section.  Some of these constraints are regulative in nature: they provide requirements for the math and therefore cannot be derived from mathematics alone. Our goal is to show how these constraints follow logically from the physical conditions, how they are encoded in the math, and how they recover well-known structures.
%to do: give a gloss on notion of regulative assumption (Kant and Peirce)

@@

First, we will show that a classical phase space arises if a system satisfies the following physical condition 
%and a series of minor regulative assumptions:
%should we indicate to the reader that the following assumption alone is insufficient to recover symplectic manifold, but it is the central assumption.

@@

Let $\mathcal{C}$ be the state space for the whole system, so that any $c \in \mathcal{C}$ represents a particular state of the whole system. Given infinitesimal reducibility, a classical system consists of a collection of classical particles. Let $\mathcal{S}$ be the state space for the particles, so that each point $s \in \mathcal{S}$ describes a possible state of a classical particle. Then for each system state $c \in \mathcal{C}$ there exists a unique distribution $\rho_c : \mathcal{S} \to \mathbb{R} $ that describes the state of $c$'s parts. That is, the state of the whole system is a distribution over its infinitesimal parts. This distribution describes the proportion of classical particles in any particular state from $\mathcal{S}$.
%ref 1: worth clarifying relationship between particle statespace S and system state space C. Is there an S_i for each i-th particle? And explain why we are not focusing on the statespace of the many particle system, e.g. a cartesian product of state spaces for each particle

@@

When the system is characterized by continuous quantities, $\mathcal{S}$ is a manifold. For $U \subseteq \mathcal{S}$, we define $\mu(U) := \int_U \rho_c(s) d\mathcal{S}$ to measure the number of particles in region $U$. \textit{State variables} $\xi^a$ are any set of quantities that fully identify states, so that $\xi$: $\mathcal{S} \to \mathbb{R}^n$ is an invertible function between each state and the numeric values of the variables.\footnote{In the language of manifolds, these $\xi^a$ are called `coordinates' and $\xi$ is a coordinate map. We will have multiple manifolds, so to avoid confusion, we will call the coordinates of the state space `state variables'. As usual with coordinate charts, $\xi$ is properly defined on some $U \subseteq \mathcal{S}$, and $\xi^{-1}$ is defined on $\xi(U) \subseteq \mathbb{R}^n$. We will neglect these subtleties in what follows.} The map $\xi$ represents a state $s \in \mathcal{S}$ by a set of state variables $\xi^a \in \mathbb{R}^n$, such that $\xi (s) = \xi^a$. To represent the distribution $\rho_c$ in terms of state variables, we define a density $\rho_{c, \xi}$: $\mathbb{R}^n \to \mathbb{R}$ by $\rho_{c, \xi} \equiv (\rho_c \circ \xi^{-1})$. This density plays a crucial role throughout our argument.

Since $\rho_c$: $\mathcal{S} \to \mathbb{R}$ is a function of the particle state $s$, we assume that its value is independent of the choice of state variables $\xi^a$ (i.e. the choice of state variable map $\xi$). If this assumption did not hold, then we could not model the system in the first place using state variables, and this would preclude many standard representational tools. Hence, the density $\rho_{c, \xi}$ must transform as a scalar under state variable changes.\footnote{For two state variable maps $\xi$ and $\hat{\xi}$, we have $\rho_c(s) = (\rho_c \circ \xi^{-1} \circ \xi)(s) = (\rho_c \circ \hat{\xi}^{-1} \circ \hat{\xi})(s)$, which entails that $\rho_c(s) = \rho_{c, \xi} (\xi^a) = \rho_{c, \hat{\xi}} (\hat{\xi}^b)$. Here, $\xi(s) \equiv \xi^a$ and $\hat{\xi}(s) \equiv \hat{\xi}^b$.} Likewise, we assume that the integral $\int_U \rho_c(s) d\mathcal{S} $ is independent of the choice of state variables, so $\int_{\xi(U)} \rho_{c, \xi} (\xi^a) d \xi^a = \int_{\hat{\xi}(U)} \rho_{c, \hat{\xi}} (\hat{\xi}^b) d \hat{\xi}^b$.\footnote{Here, we assume that $\xi$ and $\hat{\xi}$ map the background measure on $\mathcal{S}$ to the standard Lebesgue measure on $\mathbb{R}^n$. In order to match integrands below, we also assume that $\xi$ and $\hat{\xi}$ agree on their orientation. These are regulative assumptions so that we can model the system using coordinates on $\mathbb{R}^n$. They amount to assuming that the particle state space $\mathcal{S}$ has a volume form.} By the change of variables formula, we also know that $\int_{\hat{\xi}(U)} \rho_{c, \hat{\xi}} (\hat{\xi}^b) d \hat{\xi}^b= \int_{\xi(U)} \rho_{c, \hat{\xi}} (\hat{\xi}^b) \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right|  d \xi^a$.\footnote{The change of variables transformation from $\xi^a$ to $\hat{\xi}^b$ is given by $\phi \equiv \hat{\xi} \circ \xi^{-1}$: $\mathbb{R}^n \to \mathbb{R}^n $. For convenience, we denote $\rho_{c, \hat{\xi}} (\phi(\xi^a))$ by $\rho_{c, \hat{\xi}} (\hat{\xi}^b)$. Likewise, in the numerator of the Jacobian, we denote $\partial \phi$ by $\partial \hat{\xi}^b$, where $\hat{\xi}^b$ is here understood as a function of $\xi^a$.} Here, $ \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right|$ is the determinant of the Jacobian for this change of state variables. Matching integrands from the preceding two equations, we have $\rho_{c, \xi} (\xi^a) = \rho_{c, \hat{\xi}} (\hat{\xi}^b) \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right|$. From the scalar transformation property, $\rho_{c, \xi} (\xi^a) = \rho_{c, \hat{\xi}} (\hat{\xi}^b)$. Hence, we see that $\rho_{c, \hat{\xi}} (\hat{\xi}^b) =  \rho_{c, \hat{\xi}} (\hat{\xi}^b) \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right|$, entailing that the Jacobian determinant must equal one. By dimensional homogeneity, it must also be dimensionless.
